---
title: "Divvy Cyclistic Analysis - 12 months"
author: "Peter A. Blackwell"
date: "2023-02-18"
output: 
  html_document:
  anchor_sections : TRUE
  style: icon
  depth: 2
---

***

This analysis is based on the Divvy case study 
"Sophisticated, Clear, and Polished’: Divvy and Data Visualization" <br> 
written by Kevin Hartman 
( found here: https://artscience.blog/home/divvy-dataviz-case-study ). 

The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question of the case study: 

#### “In what ways do members and casual riders use Divvy bikes differently?”

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## SETUP STEPS
##### Install required packages -
##### 1) tidyverse for data import and wrangling
##### 2) lubridate for date functions
##### 3) ggplot for visualization
##### 4) dplyr for analysis
 
```{r}
library(tidyverse)  
library(lubridate)  
library(ggplot2)
library(dplyr)
```

## STEP 1: COLLECT DATA
1. Upload Divvy datasets (csv files) here - using directory 'Cyclistic - data'
```{r}
library(here)
month_1 <- read_csv(here("Cyclistic - data","202201-divvy-tripdata.csv"))
month_2 <- read_csv(here("Cyclistic - data","202202-divvy-tripdata.csv"))
month_3 <- read_csv(here("Cyclistic - data","202203-divvy-tripdata.csv"))
month_4 <- read_csv(here("Cyclistic - data","202204-divvy-tripdata.csv"))
month_5 <- read_csv(here("Cyclistic - data","202205-divvy-tripdata.csv"))
month_6 <- read_csv(here("Cyclistic - data","202206-divvy-tripdata.csv"))
month_7 <- read_csv(here("Cyclistic - data","202207-divvy-tripdata.csv"))
month_8 <- read_csv(here("Cyclistic - data","202208-divvy-tripdata.csv"))
month_9 <- read_csv(here("Cyclistic - data","202209-divvy-publictripdata.csv"))
month_10 <- read_csv(here("Cyclistic - data","202210-divvy-tripdata.csv"))
month_11 <- read_csv(here("Cyclistic - data","202211-divvy-tripdata.csv"))
month_12 <- read_csv(here("Cyclistic - data","202212-divvy-tripdata.csv"))

```

2. Stack individual months' data frames into one big data frame
```{r}
all_trips <- bind_rows(month_1, month_2, month_3, month_4, month_5, month_6
                       ,month_7, month_8, month_9, month_10, month_11, month_12)
```

## STEP 2: CLEAN DATA (PART 1)
1.  Remove columns not needed for analysis
    (station id columns, longitude/latitude columns)
```{r}    
all_trips<- all_trips %>%
  select(-c(start_station_id,end_station_id:end_lng))
```
2. Remove 'na' values - this will put into a new data frame 'all_data_trips_v2'
```{r}
all_trips_v2 <- na.omit(all_trips)
```
3. Run a count of 'member_casual' on the cleaned up data frame
```{r}
table(all_trips_v2$member_casual)
```


4. Add custom columns to the all_trips_v2 dataset based off the 'started_at' column - 
   
   extract the date without the timestamp, and month, day, year
   
   Also get the day of week in number format
```{r}
all_trips_v2$date <- as.Date(all_trips_v2$started_at) #The default format is yyyy-mm-dd
all_trips_v2$month <- format(as.Date(all_trips_v2$date), "%b")
all_trips_v2$day <- format(as.Date(all_trips_v2$date), "%d")
all_trips_v2$year <- format(as.Date(all_trips_v2$date), "%Y")
all_trips_v2$day_of_week <- format(as.Date(all_trips_v2$date), "%A")
```

5. Add a "ride_length" calculation to all_trips (in seconds)

   (reference - https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html)
```{r}
all_trips_v2$ride_length <- difftime(all_trips_v2$ended_at,all_trips_v2$started_at)
```

## STEP 3: INSPECT THE DATA
1. Inspect the structure of the columns
```{r}
str(all_trips_v2)
```

2.  Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.numeric(all_trips_v2$ride_length)
all_trips_v2$ride_length <- as.numeric(as.character(all_trips_v2$ride_length))
is.numeric(all_trips_v2$ride_length)
```
## STEP 2: CLEAN DATA (PART 2)
1. Remove "bad" data

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for   quality by Divvy or ride_length was negative.

We will create a new version of the dataframe (_clean) since data is being removed

(reference - https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/ )
```{r}
   all_trips_clean <- all_trips_v2[!(all_trips_v2$start_station_name == "HQ QR" |          all_trips_v2$ride_length<0),]
summary(all_trips_clean)
summary(all_trips_v2$ride_length)
# Run summary calculations on the ride length field for display
aggregate(all_trips_clean$ride_length ~ all_trips_clean$member_casual, FUN = mean)
aggregate(all_trips_clean$ride_length ~ all_trips_clean$member_casual, FUN = median)
aggregate(all_trips_clean$ride_length ~ all_trips_clean$member_casual, FUN = max)
aggregate(all_trips_clean$ride_length ~ all_trips_clean$member_casual, FUN = min)
```

2. Re-order the day of the week so it is American Sunday->Saturday order
```{r}
all_trips_clean$day_of_week <- ordered(all_trips_clean$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
3. Make sure the months are in Jan - Dec. order
```{r}
all_trips_clean$month <- ordered(all_trips_clean$month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul","Aug","Sep","Oct","Nov","Dec"))
```


## STEP 5: ANALYZE THE DATA
##### Note: We will use median rather than average(mean) on analyses and
##### plots to better flatten the extremes in values

1. Select number of rides, Median duration by customer type and day
```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),median_duration = median(ride_length))
```


2. Select number of rides, Median duration by customer type and month
```{r rows.print=40}
all_trips_clean %>%
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n(),median_duration = median(ride_length)) %>% 
  print(n=40)
```

## STEP 6: Charts
1. Pie chart summary comparison of rider type for last 12 months
```{r}
all_trips_clean %>%
  group_by(member_casual) %>%
  summarise(number_of_rides = n()) %>% 
  ggplot(aes(x = "", y = number_of_rides,fill=member_casual )) +
    geom_bar(stat="identity", width=1) +
    coord_polar("y", start=0) +
  geom_text(aes(label = paste0(round(number_of_rides / sum(number_of_rides) * 100, 1), "%")), 
            position = position_stack(vjust=0.5)) +
  labs(x = NULL, y = NULL, fill = NULL) +
  labs(title = "Total Rides from Divvy members exceeds casual riders", 
       subtitle = "Information reflects recording from year 2022", 
       caption = "Data collected from from Divvy from Jan. 2022 - Dec. 2022") +
  theme_void()
```

2. Analyze ridership data by type -weekday, number of riders
```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Comparison by weekday and ride time of Casual and customer riders") +
  xlab("Day of week") +
  ylab("Number of rides") +
  labs(subtitle = "Members take more rides on weekdays",
    caption = "Data collected from from Divvy from Jan. 2022 - Dec. 2022") +
  scale_y_continuous(breaks=seq(00000,450000,by=50000))
```

3. Compare customer type over last 12 months by duration of ride
```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()
            ,median_duration = median(ride_length)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = median_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Comparison by month and ride time of Casual and customer riders") +
  xlab("Month") +
  ylab("Median Length of ride (in seconds)") +
  labs(subtitle = "Casual riders take longer rides in warmer months",
    caption = "Data collected from from Divvy from Jan. 2022 - Dec. 2022") 
```

4. Analyze ridership data by type and month - number of riders
```{r}
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Comparison by month of Casual and customer riders") +
  xlab("Month") +
  ylab("Number of rides per month") +
  labs(subtitle = "Members take more rides in warmer months",
    caption = "Data collected from from Divvy from Jan. 2022 - Dec. 2022") +
  scale_y_continuous(breaks=seq(0,350000,by=50000))
```

4. Line chart comparison of duration of ride over last 12 month by customer type 
```{r}
library(ggrepel)
# Another comparison of rider type
all_trips_clean %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,median_duration = median(ride_length)) %>%
  mutate(Sat_value = if_else(weekday == 'Sat', member_casual,NULL)) %>%  
  arrange(member_casual, weekday)  %>% 
  ggplot( aes(x=weekday, y=median_duration, color=member_casual, group=member_casual)) +
  geom_line(size=1.25) +
  geom_text(aes(label = Sat_value),hjust="left") +
  ggtitle("Casual bike riders take longer rides than members") +
  theme(legend.position = "none") +
  xlab("Day of week") +
  ylab("Median length of ride in seconds") +
  labs(subtitle = "Difference is consistent on weekdays and weekends",
    caption = "Data collected from from Divvy from Jan. 2022 - Dec. 2022") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
```


## STEP 7: Output Files
##### Create a csv file that we will visualize in Excel, Tableau, or presentation software
```{r}
counts <- aggregate(all_trips_clean$ride_length ~ all_trips_clean$member_casual + all_trips_clean$day_of_week, FUN = median)
write.csv(counts, file = "C:/Users/peter/Documents/Google Data Analytics learning/Google Data Analysis docs - Part 8/Cyclistic - data/avg_ride_length.csv")
```
## STEP 8: RECOMMENDATIONS
- Management should look to appeal to casual riders via weekend campaigns, as the number of casual riders is 20% higher than on weekdays
- Casual riders take longer rides. Look for ways to appeal to riders who take longer trips in general
  (targeted discounts, etc.)
- A discount could be offered for casual summer riders to extend a yearly membership to colder winter months

```{r pressure, echo=FALSE}

```

